---
title: "Practice data.table and ggplot2"
author: 
  - name: "dr.sc. Paula Štancl"
    affiliation: "Kuzman Consulting d.o.o"
  - name: "dr.sc. Andrea Gelemanović"
    affiliation: "MedILS, UNIST"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 5
    code-fold: false
    fig-align: center
    df-print: paged
    code-summary: "Show code"
    code-line-numbers: false
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

## Library

In the first code chunk, load the **`data.table`** and **`ggplot2`** libraries --- these will be essential for data manipulation and visualization throughout the analysis.

```{r}
# Import the libraries
library(data.table)
library(ggplot2)
```

# Data set for analysis

We will analyze the [HeartDisease](https://vincentarelbundock.github.io/Rdatasets/csv/CardioDataSets/heartdisease_tbl_df.csv) dataset to explore patterns and potential risk factors associated with heart conditions. As a first step, we'll use the `fread()` function from the `data.table` package to efficiently import the dataset directly from an online source.

```{r, eval=FALSE}
heart_disease <- fread("")

```

# Tasks

1.  How many patients are in your table?

```{r}
    # Hint: Think about how to count the number of observations (rows) in a data.table.
```

2.  How many patients are *Asymptomatic* based on *ChestPain*?

```{r}
    # Hint: Use data.table syntax to filter rows and count matching cases.
```

3.  Which *ChestPain* types are more common among those with *HeartDisease*?

```{r}
# Hint: Filter by HeartDisease and summarize counts using data.table.
```

4.  Visualize the relationship between *Cholesterol* and *BP* for both sexes.

```{r}
    # Hint: Use ggplot2 (or ggpubr if preferred) to create a scatter plot grouped by sex.
```

5.  Plot the distribution of *BP* colored by *HeartDisease*.

```{r}
    # Hint: Histograms, density plots, or boxplots are all good options for showing distributions—choose one or more.

```

6.  Which female patient has the highest **MaximumHR**? What about the male patients?

```{r}
    # Hint: Use subsetting by sex and then find the row(s) with the maximum value of MaximumHR OR use by argument to get answer in one code.

```

7.  Calculate the mean value of all numeric columns.

```{r}
# Hint: Use data.table column operation syntax

```


8.  Classify patients as Young (\<35) or Old (≥35) and count them by **HeartDisease** status.

```{r}
# Hint: Use `ifelse()` inside data.table to create a new column with age group labels.

```

9.  Visualize the count of young and old patients by **HeartDisease** status.

```{r}
# Hint: Compare `geom_bar()` vs. `geom_col()`—which one fits your data format?

```
